---
title: "basic_function_in_R"
author: "Maaz AHmad"
format: html
---

# Basic function in R

In this Document, we will discuss basic function of R which are very important for every task that you performed in Rstudio.

-   Data types

-   Data structure

-   Playing with data_frame

First we will start with data types, there are four data types

## Numeric

In R, the numeric data type refers to real numbers, which can be either positive or negative and can have decimal places.

```{r}
#| echo: true
#| message: true
#| warning: true
a<-2
print(a) #to run the command press ctrl+enter
b<-3.4
print(b)
c<-2/4 #/ sign shows division
print(c)
```

To find the the data type of 2, the following command will use;

```{r}
class(a)
class(b)
class(c)
```

## Integer

Integers are whole numbers (without decimal places). In R, you can explicitly define a variable as an integer by adding an `L` after the number.

```{r}
d<- 5L
print(d)
```

Adding the `L` tells R that this value is an integer rather than a numeric (which is typically a floating-point number).

```{r}
class(d)
```

You can convert a numeric value to an integer using the `as.integer()` function.

```{r}
e<-3.4
d<-as.integer(e)
class(d)
```

Note that when converting from numeric to integer, R shorten the decimal part.

```{r}
f<-2L
g<-as.numeric(f)
class(g)
```

## Logical

The logical data type represents `TRUE` or `FALSE` values. Logical values are particularly useful when performing comparisons, controlling the flow of a program, or subsetting data frames.

```{r}
h<-2<3
print(h)
```

```{r}
j<-4>9
print(j)
```

```{r}
k<-101==101
print(k)
```

```{r}
l<-50!=50
print(l)
```

### Logical Operations

You can perform logical operations such as AND (`&`), OR (`|`), and NOT (`!`).

```{r}
m<-TRUE
n<- FALSE
#AND
print(m&n)
#OR
print(m|n)
#NOT
print(!m&n)
print(m&!n)
print(!m)
```

## Character

The character data type is used to store text or strings. Character values must be enclosed in either single (`'`) or double (`"`) quotation marks.

```{r}
test1<-"you can do it"
print(test1)
```

```{r}
class(test1)
```

Characters can be concatenated using the `paste()` function

```{r}
first<-"welocme"
second<-"to"
third<-"learning"
fourth<-"R language"
paste(first,second,third,fourth)
```

`paste0()` is a special version of `paste()` that does not automatically insert a space between each character being concatenated.

```{r}
paste0(first,second,third,fourth)
```

```{r}
x <- "I am learning R" == "IamlearningR"
y <- 100/2 == 50
z <- 19 != 18

print(x & z)     # Your written response here
print(x & y)      # Your written response here
print(y | z)      # Your written response here
print(x | y)      # Your written response here
print(x & y & z)  # Your written response here
print(x | y | z)  # Your written response here
print(x & !y & z) # Your written response here
print(x | !z)     # Your written response here
print(x & !z)     # Your written response here
print(x & !y)
```

# Data Structures

## Vectors

A vector is a sequence of elements of the same data type. It can be a sequence of numbers, characters, or logical values. Vectors are created using the `c()` function.

```{r}
#Numeric vector
num <- c(11,12,13,14,15)
#character vector
char <- c("hello","How","are","you")
#logical vector
logi_vec<- c(TRUE,FALSE,TRUE)

```

Vectors are one of the most commonly used data structures in R. You can access elements of a vector using square brackets.

```{r}
num[2:4]
num[c(2,4)]

char[2:3]
char[c(2:4)]

logi_vec[2]
logi_vec[2:3]
```

## Matrices

A matrix is a two-dimensional collection of elements of the same data type. You can think of it as a table with rows and columns. You can create a matrix using the `matrix()` function.

```{r}
?matrix
mt1<-matrix(1:15, nrow = 3,ncol = 5)# bydefault byrow=FASE

mt2<-matrix(16:30,nrow = 3,ncol = 5,byrow = TRUE)#byrow arrange number sequence in matrix in row-wise
t(mt2) #t function is transpose which means it interchange rows into columns and vice versa. 

```

### **Combine Matrices**

In order to combine two matrices, the **cbind()** function is used which akes two matrices as arguments and produces their combination.When combining matrices, make sure the they have the same number of rows, otherwise an exception is thrown.

```{r}
?cbind
cbind(mt1,mt2)# combine both matrix on the basis of column

rbind(mt1,mt2)# combine both matrix on the basis of rows
```

You can access elements of a matrix using row and column indices.

```{r}
#accessing thge third row
mt1[3,]
#accessing the third column
mt1[,3]
#accessing the element in second row and third column
mt1[2,3]
```

### **Deconstruction**

This concept allows you to break down the matrix into its original vector, which can come handy in certain situations.

```{r}
c(mt1) # c function break down matrix into its original vector
```

## Arrays

An array is similar to a matrix but can have more than two dimensions. You can create an array using the `array()` function.

```{r}
?array
ay1<- array(1:20,dim = c(5,2,2))
print(ay1)
```

Here, the `dim` argument specifies the dimensions of the array: 3 rows, 2 columns, and 2 layers.

You can access elements of an array using row, column, and layer indices.

```{r}
#accessing the second row
ay1[2,,]
#accessing the elemnet in the first row and first column
ay1[1,1,]
#accesssing the element in the fourth row,second column and first layer
ay1[4,2,1]
```

## Lists

A list is a collection of elements that can be of different data types. Lists are very versatile because they can contain numbers, strings, vectors, or even other lists.

```{r}
lt1<-list(name="john",age=12,marks=c(12,16,18,10))
```

```{r}
# Accessing elements of a list
lt1$name
lt1[["age"]]
```
